---
title: "HomeWork2"
author: "Yifan Jian"
date: "2023/10/2"
output:
  html_document: default
  pdf_document: default
---

###import the library

```{r}
library('lme4')
library('ggplot2')
```

###Load data
```{r}
lead_data <- read.table(file='D:\\Graduation\\lead.dat', header=FALSE, sep = "")
colnames(lead_data) = c('child_id' , 'age_indicator', 'gender_indicator', 'week', 'lead_level', 'treatment')
head(lead_data)
```

###a) 
```{r}

# Rename the treatment levels for better readability
lead_data$treatment <- factor(lead_data$treatment, levels = c(0, 1, 2), labels = c("Placebo", "Low Dose", "High Dose"))

# Plot the blood lead levels over time for each child, colored by treatment group
ggplot(lead_data, aes(x = week, y = lead_level, group = child_id, color = treatment)) +
  geom_line(alpha = 0.5) +
  labs(title = "Blood Lead Levels Over Time by Treatment Group",
       x = "Week",
       y = "Blood Lead Level") +
  theme_minimal()
```
_From the plot, we can make some observations:_

**The trajectories of blood lead levels for individual children seem to vary, with some showing increases, some showing decreases, and others remaining relatively stable.

**There is a considerable overlap between the trajectories of the three treatment groups, making it challenging to discern clear patterns at this preliminary stage.

**While many trajectories seem to be approximately linear, there are also some that exhibit more complex patterns.

###b)
we fit the model with each a assumption
```{r}

# Fit the models
# Model 1: Same for all treatments
model1 <- lmer(lead_level ~ week + (1 + week | child_id), data = lead_data, REML = TRUE)

# Model 2: Different fluctuations, same intercepts/slopes
model2 <- lmer(lead_level ~ week + treatment + (1 + week | child_id) + (1 | treatment:child_id), data = lead_data, REML = TRUE)

# Model 3: Same fluctuations, different intercepts/slopes
model3 <- lmer(lead_level ~ week + treatment + (1 + week | treatment:child_id), data = lead_data, REML = TRUE)

# Model 4: Different for all
model4 <- lmer(lead_level ~ week + treatment + (1 + week | treatment:child_id) + (1 | treatment:child_id), data = lead_data, REML = TRUE)

# Compare models using AIC and BIC
aic_bic <- data.frame(Model = c("Model 1", "Model 2", "Model 3", "Model 4"),
                      AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
                      BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4)))

print(aic_bic)
```
###c)
From inspection of AIC and BIC for each model fit,we conlude that Model 3 have the lowest AIC and BIC. therefore, Magnitude of within-child fluctuations in lead levels is the same under all treatments but the way in which child-specific intercepts and slopes vary/co-vary are possibly different will be the assumption we prefer. 

###d)
```{r}
# Load necessary libraries
library(lme4)
library(aod)  # For wald.test function

# Fit the preferred Model 3 with interaction term
model3_interact <- lmer(lead_level ~ week * treatment + (1 + week | treatment:child_id), data = lead_data, REML = TRUE)

# Identify the indices of the interaction terms in the coefficient vector
interaction_indices <- which(grepl("week:treatment", names(fixef(model3_interact))))

# Conduct Wald test for the interaction terms using identified indices
result <- wald.test(b = fixef(model3_interact), Sigma = vcov(model3_interact), Terms = interaction_indices)

# Print the Wald test result
print(result)

```
